Fractional Types

### 2013-02-17 frac.tex
changes to section 1-3

We are witnessing a convergence of ideas from several distinct research
communities (physics, mathematics, and computer science) towards replacing
\emph{equalities} by \emph{isomorphisms}. The combined programme originating
from these research areas has sparked a huge amount of research that unveiled
new and surprising connections between geometry, algebra, logic, and
computation (see~\cite{baez2011physics} for an overview).

In the physics community, Landauer~\cite{Landauer,Landauer:1961},
Feynman~\cite{springerlink:10.1007/BF02650179}, and others have interpreted
the laws of physics as fundamentally related to computation. In more detail,
the great majority of the laws of physics are formulated as equalities
between different physical observables. For instance, Newton's second law of
classical mechanics equates the force acting on a system to its rate of
change of momentum. The insight of Landauer, Feynman, etc. is that
\emph{different} physical observables should not be related by an
\emph{equality} but more appropriately by an \emph{isomorphism} that
witnesses, explains, and models the process of transforming one observable to
the other.

In the mathematics and logic community, Martin-L\"of developed an extension
of the simply typed $\lambda$-calculus originally intended to provide a
rigorous framework for constructive
mathematics~\cite{citeulike:7374951}. This theory has been further extended
with \emph{identity types} representing the proposition that two terms are
``equal.'' (See~\cite{warren,streicher} for a survey.) Briefly speaking,
given two terms $a$ and $b$ of the same type $A$, one forms the type
$\texttt{Id}_A(a,b)$ representing the proposition that $a$ and $b$ are equal:
in other words, a term of type $\texttt{Id}_A(a,b)$ witnesses, explains, and
models the process of transforming $a$ to $b$ and vice-versa.

@@

A major open problem remains, however. This concerns a higher-order extension
of {{Pi}}. This extension is of fundamental importance in all the
originating research areas. In physics, it allows the process itself or the
observer to be treated as ``data'' that can be transformed by higher-level
processes or observed by meta-level observers. In mathematics and logic, it
allows the equivalence between different proofs of type $\texttt{Id}_A(a,b)$
to be itself expressed as an isomorphism as a higher type
$\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
higher-order types allow code to abstract over other code fragments as well
as the manipulation of code as data and data as code which is central to many
algorithms and application domains.

@@

We review our language {{Pi}} providing the necessary background and
context for our higher-order extension.\footnote{The presentation in this
  section focuses on the simplest version of {{Pi}}. Other versions
  include the empty type, recursive types, and trace operators but these
  extensions are orthogonal to the higher-order extension emphasized in this
  paper.} The terms of {{Pi}} are not classical values and functions;
rather, the terms are isomorphism witnesses.  In other words, the terms of
{{Pi}} are proofs that certain ``shapes of values'' are isomorphic.
And, in classical Curry-Howard fashion, our operational semantics shows how
these proofs can be directly interpreted as actions on ordinary values which
effect this shape transformation. Of course, ``shapes of values'' are very
familiar already: they are usually called \emph{types}.  But usually one
designs a type system as a method of classifying terms, with the eventual
purpose to show that certain properties of well-typed terms hold, such as
safety.  Our approach is different: we start from a type system, and then
present a term language which naturally inhabits these types, along with an
appropriate operational semantics.

@@

\paragraph*{Isomorphisms.} The terms of {{Pi}} are witnesses to type
isomorphisms.  They have (iso) types {{b <-> b}}.  
Specifically, they are witnesses to the following type isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
\noindent Each line of the above table introduces a pair of dual
constants~\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness
the type isomorphism in the middle.  These are the base (non-reducible) terms
of the second, principal level of {{Pi}}. Note how the above has two
readings: first as a set of typing relations for a set of constants. Second,
if these axioms are seen as universally quantified, orientable statements,
they also induce transformations of the (traditional) values. The
(categorical) intuition here is that these axioms have computational content
because they witness isomorphisms rather than merely stating an extensional
equality.

@@

%% AMR:
%% \jacques{Since we have Agda code for the above, should we refer to it
%% here, and make it available?}
%% That is a great idea but it would take a significant amount of time to
%% clean up the code and make it presentable. If someone is willing to do
%% that, then we should definitely add a reference to the code and make it
%% available.

The language presented above, at the type level, models a commutative ringoid
where the multiplicative structure forms a commutative monoid, but the
additive structure is just a commutative magma.  Note that the version of
{{Pi}} that include the empty type with its usual laws exactly captures, at
the type level, the notion of a \emph{semiring} (occasionally called a
\emph{rig}) where we replace equality by isomorphism.  This can also be also
compared to \emph{bimonoidal categories}.  As is usual with type theories,
the coherence conditions (pentagonal and triangle identity) are not explicit,
but would correspond to certain identity types being trivial.

@@

% AMR: a function a -o b is modeled by (1/a,b) not (a,b) 
% \roshan{It is unclear how fractions help differentiate between inputs
%   and outputs. On first reading this seemed promising, but nothing
%   in the rest of the paper substantiates this point of view.}
%

% AMR: relations are indeed only presented as a way to explain the semantics
% \roshan{I think we should arrive at relational programming as a means
%   of giving semantics to fractionals, rather than the other way
%   round. To show that relational programming is reversible, one does
%   not need to resort to fractionals. Everyone understand that
%   relations can be reversed and none of this setup with {{Pi}} is
%   required for that. Our key point should be to communicate that
%   fractionals have a clean semantics in the setting of relational
%   programming rather than the other way round.}

@@

\begin{definition}[Syntax of {{PiTF}}]
\label{def:PiT}
We extend Def.~\ref{def:Pi} with:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v
%
% base iso ::= ... | eta*_b | eps*_b
where:
{{eta*_b : 1 <-> 1/b * b : eps*_b}}
\end{definition}

@@

\begin{definition}[Relational semantics]
\label{def:relational-PiTF}
Each combinator {{c : b1 <-> b2}} in {{PiTF}} denotes a relation
as specified below. 
%subcode{proof} include main
%@ ~
%@@ (left v) ~~swap+~~ (right v)
%
%@ ~
%@@ (right v) ~~swap+~~ (left v)
%----
%@ ~
%@@ (left v) ~~assocl+~~ (left (left v))
%
%@ ~
%@@ (right (left v)) ~~assocl+~~ (left (right v))
%----
%@ ~
%@@ (right (right v)) ~~assocl+~~ (right v)
%
%@ ~
%@@ (left (left v)) ~~assocr+~~ (left v)
%----
%@ ~
%@@ (left (right v)) ~~assocr+~~ (right (left v))
%
%@ ~
%@@ (right v) ~~assocr+~~ (right (right v))
%----
%@ ~
%@@ ((), v) ~~unite~~ v
%
%@ ~
%@@ v ~~uniti~~ ((), v)
%----
%@ ~
%@@ (v1, v2) ~~swap*~~ (v2, v1)
%----
%@ ~
%@@ (v1, (v2, v3)) ~~assocl*~~ ((v1, v2), v3)
%
%@ ~
%@@ ((v1, v2), v3) ~~assocr*~~ (v1, (v2, v3))
%----
%@ ~
%@@ (left v1, v3) ~~dist~~ (left (v1, v3))
%
%@ ~
%@@ (right v2, v3) ~~dist~~ (right (v2, v3))
%----
%@ ~
%@@ (left (v1, v3)) ~~factor~~ (left v1, v3)
%
%@ ~
%@@ (right (v2, v3)) ~~factor~~ (right v2, v3)
%----
%@ ~
%@@ v ~~id~~ v
%
%@ v' ~~c~~ v
%@@ v ~~(sym c)~~ v'
%
%@ v1 ~~c1~~ v2
%@ v2 ~~c2~~ v3
%@@ v1 ~~(c1(;)c2)~~ v3
%----
%@ v ~~c1~~ v'
%@@ (left v) ~~(c1 (+) c2)~~ (left v')
%
%@ v ~~c2~~ v'
%@@ (right v) ~~(c1 (+) c2)~~ (right v')
%----
%@ v1 ~~c1~~ v1'
%@ v2 ~~c2~~ v2'
%@@ (v1, v2) ~~(c1 (*) c2)~~ (v1', v2')
%----
%@ v in [[ b ]]  
%@@ () ~~eta*_b~~ (1/v,v)
%
%@ ~
%@@ (1/v,v) ~~eps*_b~~ ()
\end{definition}

>>>>

We are witnessing a convergence of ideas from several distinct research
communities (physics, mathematics, and computer science) towards replacing
\emph{equalities} by \emph{isomorphisms}. The combined programme 
has sparked a huge amount of research that unveiled
new and surprising connections between geometry, algebra, logic, and
computation (see~\cite{baez2011physics} for an overview).

In the physics community, Landauer~\cite{Landauer:1961,Landauer},
Feynman~\cite{springerlink:10.1007/BF02650179}, and others have interpreted
the laws of physics as fundamentally related to computation. In more detail,
the great majority of the laws of physics are formulated as equalities
between different physical observables, such as Newton's second law of
classical mechanics ($F=\dot{p}$).
The insight of Landauer, Feynman, etc. is that
\emph{different} physical observables should not be related by an
\emph{equality} but more appropriately by an \emph{isomorphism} that
witnesses, explains, and models the process of transforming one observable to
the other.

In the mathematics and logic community, Martin-L\"of developed an extension
of the simply typed $\lambda$-calculus originally intended to provide a
rigorous framework for constructive
mathematics~\cite{citeulike:7374951}. This theory has been further extended
with \emph{identity types} representing the proposition that two terms are
``equal.'' (See~\cite{streicher,warren} for a survey.) Briefly speaking,
given two terms $a$ and $b$ of the same type $A$, one forms the type
$\texttt{Id}_A(a,b)$ representing the proposition that $a$ and $b$ are equal:
in other words, a term of type $\texttt{Id}_A(a,b)$ witnesses, explains, and
models the process of transforming $a$ to $b$ and vice-versa.

@@

A major open problem remains, however: a higher-order extension
of {{Pi}}. This extension is of fundamental importance in all the
originating research areas. In physics, it allows the process itself or the
observer to be treated as ``data'' that can be transformed by higher-level
processes or observed by meta-level observers. In mathematics and logic, it
allows the equivalence between different proofs of type $\texttt{Id}_A(a,b)$
to itself be expressed as an isomorphism (of a higher type)
$\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
higher-order types allow code to abstract over other code fragments as well
as the manipulation of code as data and data as code.

@@

We review our language {{Pi}} providing the necessary background and
context for our higher-order extension.\footnote{The presentation in this
  section focuses on the simplest version of {{Pi}}. Other versions
  include the empty type, recursive types, and trace operators but these
  extensions are orthogonal to the higher-order extension emphasized in this
  paper.} The terms of {{Pi}} are not classical values and functions;
rather, the terms are isomorphism witnesses.  In other words, the terms of
{{Pi}} are proofs that certain ``shapes of values'' are isomorphic.
And, in classical Curry-Howard fashion, our operational semantics shows how
these proofs can be directly interpreted as actions on ordinary values which
effect this shape transformation. Of course, ``shapes of values'' are very
familiar already: they are usually called \emph{types}.  But frequently one
designs a type system as a method of classifying terms, with the eventual
purpose to show that certain properties of well-typed terms hold, such as
safety.  Our approach is different: we start from a type system, and then
present a term language which naturally inhabits these types, along with an
appropriate operational semantics.

@@

\paragraph*{Isomorphisms.} The important terms of {{Pi}} are witnesses to
(value) type isomorphisms.  They are also typed, by the shape of the (value)
type isomorphism they witness {{b <-> b}}.  Specifically, they are witnesses to
the following type isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
\noindent Each line of the above table introduces a pair of dual
constants\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness
the type isomorphism in the middle.  These are the base (non-reducible) terms
of the second, principal level of {{Pi}}. Note how the above has two
readings: first as a set of typing relations for a set of constants. Second,
if these axioms are seen as universally quantified, orientable statements,
they also induce transformations of the (traditional) values. The
(categorical) intuition here is that these axioms have computational content
because they witness isomorphisms rather than merely stating an extensional
equality.

@@

%% AMR:
%% \jacques{Since we have Agda code for the above, should we refer to it
%% here, and make it available?}
%% That is a great idea but it would take a significant amount of time to
%% clean up the code and make it presentable. If someone is willing to do
%% that, then we should definitely add a reference to the code and make it
%% available.
%% Jacques:  Right.  I'll definitely work on cleaning up the paper first,
%% then see if there is time left for the code.

The language presented above, at the type level, models a commutative ringoid
where the multiplicative structure forms a commutative monoid, but the
additive structure is just a commutative semigroup.  Note that the version of
{{Pi}} that include the empty type with its usual laws exactly captures, at
the type level, the notion of a \emph{semiring} (occasionally called a
\emph{rig}) where we replace equality by isomorphism.  This can also be also
compared to \emph{bimonoidal categories}.  As is usual with type theories,
the coherence conditions (pentagonal and triangle identity) are not explicit,
but would correspond to certain identity types being trivial.

@@

% \roshan{It is unclear how fractions help differentiate between inputs
%   and outputs. On first reading this seemed promising, but nothing
%   in the rest of the paper substantiates this point of view.}
% AMR: a function a -o b is modeled by (1/a,b) not (a,b) 
% Jacques: Exactly!  1/a is negative information is input, b is positive is
%    output.
%

% \roshan{I think we should arrive at relational programming as a means
%   of giving semantics to fractionals, rather than the other way
%   round. To show that relational programming is reversible, one does
%   not need to resort to fractionals. Everyone understand that
%   relations can be reversed and none of this setup with {{Pi}} is
%   required for that. Our key point should be to communicate that
%   fractionals have a clean semantics in the setting of relational
%   programming rather than the other way round.}
% AMR: relations are indeed only presented as a way to explain the semantics

@@

\begin{definition}[Syntax of {{PiTF}}]
\label{def:PiT}
We extend our previous definition with:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v
%
% base iso ::= ... | eta*_b | eps*_b
where:
{{eta*_b : 1 <-> 1/b * b : eps*_b}}
\end{definition}

@@

\begin{definition}[Relational semantics]
\label{def:relational-PiTF}
Each combinator {{c : b1 <-> b2}} in {{PiTF}} denotes a relation
as specified below. 
%subcode{proof} include main
%@ ~
%@@ (left v) ~~swap+~~ (right v)
%
%@ ~
%@@ (right v) ~~swap+~~ (left v)
%----
%@ ~
%@@ (left v) ~~assocl+~~ (left (left v))
%
%@ ~
%@@ (right (left v)) ~~assocl+~~ (left (right v))
%----
%@ ~
%@@ (right (right v)) ~~assocl+~~ (right v)
%
%@ ~
%@@ (left (left v)) ~~assocr+~~ (left v)
%----
%@ ~
%@@ (left (right v)) ~~assocr+~~ (right (left v))
%
%@ ~
%@@ (right v) ~~assocr+~~ (right (right v))
%----
%@ ~
%@@ ((), v) ~~unite~~ v
%
%@ ~
%@@ v ~~uniti~~ ((), v)
%----
%@ ~
%@@ (v1, v2) ~~swap*~~ (v2, v1)
%----
%@ ~
%@@ (v1, (v2, v3)) ~~assocl*~~ ((v1, v2), v3)
%
%@ ~
%@@ ((v1, v2), v3) ~~assocr*~~ (v1, (v2, v3))
%----
%@ ~
%@@ (left v1, v3) ~~dist~~ (left (v1, v3))
%
%@ ~
%@@ (right v2, v3) ~~dist~~ (right (v2, v3))
%----
%@ ~
%@@ (left (v1, v3)) ~~factor~~ (left v1, v3)
%
%@ ~
%@@ (right (v2, v3)) ~~factor~~ (right v2, v3)
%----
%@ ~
%@@ v ~~id~~ v
%
%@ v' ~~c~~ v
%@@ v ~~(sym c)~~ v'
%
%@ v1 ~~c1~~ v2
%@ v2 ~~c2~~ v3
%@@ v1 ~~(c1(;)c2)~~ v3
%----
%@ v ~~c1~~ v'
%@@ (left v) ~~(c1 (+) c2)~~ (left v')
%
%@ v ~~c2~~ v'
%@@ (right v) ~~(c1 (+) c2)~~ (right v')
%----
%@ v1 ~~c1~~ v1'
%@ v2 ~~c2~~ v2'
%@@ (v1, v2) ~~(c1 (*) c2)~~ (v1', v2')
%----
%@ ~
%@@ () ~~eta*_b~~ (1/v,v)
%
%@ ~
%@@ (1/v,v) ~~eps*_b~~ ()
\end{definition}
